Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-09-28 ms.tex
Fixing matching conditions in WKB appendix

The solution of the radiative transfer equation for resonant scattering of Lyman $\alpha$ photons is explored in spherical geometry. A monochromatic source of photons is assumed at the center of a uniform sphere of constant gas density. The intensity within the sphere and at its surface is found in the limit of large line-center optical depth using the Eddington approximation for the angular dependence. Two specific problems are studied: a steady source and an impulsive source. For the steady-state problem, the solution can be represented as a sum of two terms: a previously-known analytic solution of the homogeneous equation with the mean intensity $J=0$ at the surface, and a novel, semi-analytic solution of the homogeneous equation which enforces the correct, zero-ingoing intensity boundary conditions at the surface. This solution is compared to that of the Monte-Carlo method, which is valid at arbitrary optical depth. It is shown that the size of corrective term depends on the line center optical depth, and may explain discrepancies in previous studies that used a $J=0$ boundary condition. For the impulsive problem, the time dependence, spatial dependence, and photon frequency dependence of the diffusion equation is expressed using an eigenfunction expansion and solved numerically to characterize the escape time distribution and spectra of photons initialized at the center of the sphere. It is shown that the lowest-order eigenfrequency of the semi-analytic calculation agrees well with the decay constant of the Monte Carlo escape time distribution at line-center optical depths above ${\sim} 10^7$. A fitting function for the distribution is found which can be sampled to accelerate Monte Carlo radiation transfer calculations at very large optical depths.

@@

In Fig. \ref{fig:steadystate}, the spectrum for $x_s=0$ and $\tau_0=10^7$ is shown for a sum up to $n=20$ and $m=500$ (labelled ``Time-integrated'') as compared with two analytic solutions: the steady state $H_0$ solution with no time dependence (``Steady State'') and the result for summing \textit{only} spatial modes in the eigenfunction expansion with the time dependence ignored (``Partial Sum''), as in Eq. \ref{eq:H0surf}. Additional spatial modes contribute to the solutions' accuracy in the core of the line. If more spatial modes are calculated, the agreement with the steady state spectrum extends further toward the line core, with an infinite number of modes required for convergence at $x_s$. If additional frequency modes are calculated, faster-oscillating terms are incorporated into the sum over $\gamma$ which create more perfect cancellations with the lower-order terms, reducing the ``ringing'' seen in the time-integrated spectrum. Extending the calculation deep into the line core by adding additional spatial modes could have an impact on the accuracy of the escape time distribution, but this would only affect the distribution at late times since photons are trapped for a longer duration before escaping. This was the motivation for choosing a comparatively low number of spatial eigenmodes with respect to the number of frequency eigenmodes calculated. %TODO: Include? As a result of this choice, we expect that our semianalytic approach will underestimate the number of escaping photon at late times as compared with the Monte Carlo.

@@

In Figure \ref{fig:escape_time} escape time distributions calculated from Eq. \ref{eq:waittime} are shown alongside Monte Carlo and the fitting function Eq. \ref{eq:fitting_function} for $\tau_0=10^6, 10^7$ with $x_s=0$. It is seen that the fitting function matches well with the late-time behavior of the eigenfunction solution, indicating that the tail of the escape time distribution can be modeled effectively by simple exponential decay. The remaining discrepancy between the tail of the distribution and the Monte Carlo data is likely due to line core scattering which is not modeled by the eigenfunction solution. A cumulative distribution function can be created from Eq. \ref{eq:fitting_function} using a simple rejection method, which allows the function to be sampled rapidly.

@@

We start with the differential equation Eq. \ref{eq:wkb_differential_eqn}. Near line center, this has two independent solutions which can be written
\be
J(n,\sigma,s) & = & a_1 \left( \frac{\sigma}{\sigma_{\rm tp}} \right)^{1/2} J_{-3/4}\left( k_c \left[ \frac{\sigma}{\sigma_{\rm tp} }\right]^{2/3} \right)
\nonumber \\ &  + &  a_2 \left( \frac{\sigma}{\sigma_{\rm tp}} \right)^{1/2} J_{3/4}\left( k_c \left[ \frac{\sigma}{\sigma_{\rm tp} }\right]^{2/3} \right),
\ee
where $J_\alpha(...)$ are Bessel functions of the first kind, $a_1$ and $a_2$ are normalization constants, $\sigma_{tp}$ is given in Eq. \ref{eq:sigma_tp}, and $k_c=3\kappa_n \Delta \sigma_{\rm tp}/2k$. The solution far from line center is 
\be
J(n,\sigma,s) & = & a_3 \left( \frac{\sigma}{\sigma_{\rm tp}} \right)^{1/6} Ai\left(k_c^{2/3} \left[ \left( \frac{\sigma}{\sigma_{\rm tp}} \right)^{2/3} - 1 \right]\right),
\ee
where $Ai(...)$ is the Airy function of the first kind. Again, $a_3$ is a normalization constant. When $\sigma_s=0$, we can enforce the discontinuity in Eq. \ref{eq:matching_condition_2} to find 
\be
a_2 & = & - \frac{\Gamma(7/4) \sigma_{\rm tp}}{ 2^{1/4} k_c^{3/4}} \frac{6^{1/2}}{8} n^2 \frac{E}{kR^3}.
\ee
The discontinuity does not set the value of $a_1$ as that solution does not have a term linear in $\sigma$ when the source is at $\sigma_s=0$. However, we may also show that the solutions near and far from $\sigma=0$ have the same form where both functions have short-wavelength behavior. This allows us to relate the normalization coefficients $a_1$ and $a_3$, as well as quantize the phase of the solutions to determine a dispersion relation. The matching conditions are
\be
a_1 & = & |a_2| \\ 
a_3 & = & |a_2| 2^{3/2} k_c^{-1/3}, \\
\ee
and the phase is set by
\be
\frac{\kappa_n \sigma_{tp} \Delta}{k} & = & 2\pi b
\ee
to match the two solutions, where $b$ is a positive integer. Since $\sigma_{tp}$ contains the eigenvalue, we can solve the expression for $\gamma_{nm}$ to find the dispersion relation in Eq. \ref{eq:gamma_nm}.

>>>>

The solution of the radiative transfer equation for resonant scattering of Lyman $\alpha$ photons is explored in spherical geometry. A monochromatic source of photons is assumed at the center of a uniform sphere of constant gas density. The intensity within the sphere and at its surface is found in the limit of large line-center optical depth using the Eddington approximation for the angular dependence. Two specific problems are studied: a steady source and an impulsive source. For the steady-state problem, the solution can be represented as a sum of two terms: a previously-known analytic solution of the homogeneous equation with the mean intensity $J=0$ at the surface, and a novel, semi-analytic solution of the homogeneous equation which enforces the correct, zero-ingoing intensity boundary conditions at the surface. This solution is compared to that of the Monte-Carlo method, which is valid at arbitrary optical depth. It is shown that the size of corrective term depends on the line center optical depth, and may explain discrepancies in previous studies that used a $J=0$ boundary condition. For the impulsive problem, the time dependence, spatial dependence, and photon frequency dependence of the diffusion equation are expressed using an eigenfunction expansion and solved numerically to characterize the escape time distribution and spectra of photons initialized at the center of the sphere. It is shown that the lowest-order eigenfrequency of the semi-analytic calculation agrees well with the decay constant of the Monte Carlo escape time distribution at line-center optical depths above ${\sim} 10^7$. A fitting function for the distribution is found which can be sampled to accelerate Monte Carlo radiation transfer calculations at very large optical depths.

@@

In Fig. \ref{fig:steadystate}, the spectrum for $x_s=0$ and $\tau_0=10^7$ is shown for a sum up to $n=20$ and $m=500$ (labelled ``Time-integrated'') as compared with two analytic solutions: the steady state $H_0$ solution with no time dependence (``Steady State'') and the result for summing \textit{only} spatial modes in the eigenfunction expansion with the time dependence ignored (``Partial Sum''), as in Eq. \ref{eq:H0surf}. Additional spatial modes contribute to the solutions' accuracy in the core of the line. If more spatial modes are calculated, the agreement with the steady state spectrum extends further toward the line core, with an infinite number of modes required for convergence at $x_s$. If additional frequency modes are calculated, faster-oscillating terms are incorporated into the sum over eigenmodes which create more perfect cancellations with the lower-order terms, reducing the ``ringing'' seen in the time-integrated spectrum. Extending the calculation deep into the line core by adding additional spatial modes could have an impact on the accuracy of the escape time distribution, but this would primarily affect the distribution at late times due to the large mean diffusion time of photons near the line core. This was the motivation for choosing a comparatively low number of spatial eigenmodes with respect to the number of frequency eigenmodes calculated. %TODO: Include? As a result of this choice, we expect that our semianalytic approach will underestimate the number of escaping photon at late times as compared with the Monte Carlo.

@@

In Figure \ref{fig:escape_time} escape time distributions calculated from Eq. \ref{eq:waittime} are shown alongside Monte Carlo and the fitting function Eq. \ref{eq:fitting_function} for $\tau_0=10^6, 10^7$ with $x_s=0$. It is seen that the fitting function matches well with the late-time behavior of the eigenfunction solution, indicating that the tail of the escape time distribution can be modeled effectively by simple exponential decay. The remaining disagreement between the tail of the distribution and the Monte Carlo data is likely due to line core scattering which is not modeled by the eigenfunction solution, but improves for larger optical depth as seen in the figure. A cumulative distribution function can be created from Eq. \ref{eq:fitting_function} using a simple rejection method, which allows the function to be sampled rapidly.

@@

We start with the differential equation Eq. \ref{eq:wkb_differential_eqn}. Near line center, this has two independent solutions which can be written
\be
J(n,\sigma,s) & = & a_1 \left( \frac{\sigma}{\sigma_{\rm tp}} \right)^{1/2} J_{-3/4}\left( k_c \left[ \frac{\sigma}{\sigma_{\rm tp} }\right]^{2/3} \right)
\nonumber \\ &  + &  a_2 \left( \frac{\sigma}{\sigma_{\rm tp}} \right)^{1/2} J_{3/4}\left( k_c \left[ \frac{\sigma}{\sigma_{\rm tp} }\right]^{2/3} \right),
\ee
where $J_\alpha(...)$ are Bessel functions of the first kind, $a_1$ and $a_2$ are normalization constants, $\sigma_{tp}$ is given in Eq. \ref{eq:sigma_tp}, and $k_c=3\kappa_n \Delta \sigma_{\rm tp}/2k$. The solution far from line center is 
\be
J(n,\sigma,s) & = & a_3 \left( \frac{\sigma}{\sigma_{\rm tp}} \right)^{1/6} Ai\left(k_c^{2/3} \left[ \left( \frac{\sigma}{\sigma_{\rm tp}} \right)^{2/3} - 1 \right]\right),
\ee
where $Ai(...)$ is the Airy function of the first kind. Again, $a_3$ is a normalization constant. When $\sigma_s=0$, we can enforce the discontinuity in Eq. \ref{eq:matching_condition_2} to find 
\be
a_2 & = & - \frac{\Gamma(7/4) \sigma_{\rm tp}}{ 2^{1/4} k_c^{3/4}} \frac{6^{1/2}}{8} n^2 \frac{E}{kR^3}.
\ee
The discontinuity does not set the value of $a_1$ as that solution does not have a term linear in $\sigma$ when the source is at $\sigma_s=0$. However, where both functions have short-wavelength behavior we may show that the solutions near and far from $\sigma=0$ have the same approximate form in terms of sines and cosines. This matched asymptotic expansion allows us to relate the normalization coefficients $a_1$ and $a_3$. The matching conditions are
\be
\frac{a_1}{a_2} & = & \frac{ \sin \left( \frac{\kappa_n \sigma_{\rm tp} \Delta}{k} + \frac{\pi}{8} \right) }{ \sin \left(  \frac{\kappa_n \sigma_{\rm tp} \Delta}{k} - \frac{\pi}{8} \right) }
\\
\frac{a_3}{a_2} & = & \frac{ k_c^{-1/3} }{ \sin \left(  \frac{\kappa_n \sigma_{\rm tp} \Delta}{k} - \frac{\pi}{8} \right) }.
\ee
By setting the to denominator of this expression equal to zero and solving it for the eigenfrequency $\gamma_{nm}$ which is contained in $\sigma_{tp}$, we find the dispersion relation in Eq. \ref{eq:gamma_nm}.
